### 2017-01-24 manuscript.rmd
typos, polishing

Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in Biology. There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity. Most published comparisons of genome function across species have, however, neglected to use methods that consider the evolutionary relationships among species. This has two potential negative impacts. First, it can lead to the wrong conclusions about the evolution of genome function. Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships. Here we examine two recently published comparative expression studies that used multiple pairwise comparison rather than phylogenetic approaches to investigate evolutionary patterns of gene expression. We find problems with the pairwise comparisons used in both studies that undermine their published conclusions. The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpretted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication. Our phylogenetic analysis finds no difference in rates following duplication and speciation. Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs. The second study claimed to find more variance in gene expression at the midpoint of development than earlier and later when comparing animals from different phyla. We instead find that changes along the branch of a single taxon, the ctenophore, were overcounted becuase they were included in each pairwise comparison. These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, and indicates that it will be critical to adopt phylogenetic comparative methods in future analyses.

@@

The focus of genomics research has quickly shifted from describing genome sequences to Functional Genomics (FG), the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation. FG, in turn, is now becoming more comparative- there is great interest in understanding how FG variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh]. Comparative Functional Genomics (CFG) analyses are also critical to transferring functional insight across species, and will grow in importance in coming years.

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Chamberlain:2012vx; @Garamszegi:2014ew; @Pagel:1999gc]. These have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG. Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on pairwise comparisons (Fig. 1A) that do not account for the fact that evolutionary relationships explain much of the structure of variation across species. This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic approaches can lead to different results in CFG. Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other character data [@Chamberlain:2012vx; @Ricklefs:1996wb], this value is apparently not well known to the FG community since the field has until recently largely focused on within-species analyses. Many initial CFG studies have also been focused on only two species at a time, where the differences between phylogenetic and pairwise approaches are not as pronounced. As the number of species considered in each CFG study grows, the shortcomings of pairwise approaches will become far more acute.

Most CFG studies to date compare gene expression across species, often between distinct tissues or developmental time points. Gene expression has been a popular target of CFG studies because, with the advent of RNA-Seq, it is the most readily collected FG trait across diverse species. Here we reconsider two recent comparative analyses of gene expression data [@KryuchkovaMostacci:2016iw; @Levin:2016bp].

Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw] integrated vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend of have more conserved attributes (expression in this case) than do paralogs [@Nehrt:2011bm]. Using pairwise comparisons (Fig. 1A), they found lower expression covariance between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture. Orthology and paralogy is annotated based on gene phylogenies, but these phylogenies are not used when making the expression comparisons.

The second CFG study we evaluate here is Levin *et al.* [@Levin:2016bp]. This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. They concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Fig. 2). This contrasts with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw]. Levin *et al.* conclude that this provides biological justification for the concept of phyla and may provide a definition of phyla. We previously described concerns with their interpretations of these results [@Hejnol:2016hw]. Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.

@@

#### Original analyses

The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). It has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is used to relate findings from well-studied genes to related genes that haven't been investigated in detail. It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it can also be considered a hypothesis about the process of evolution. Specifically, it postulates that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

Despite its importance, there have been relatively few tests of the ortolog conjecture. Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz]. Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw]. There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such study to date.

The KMRR study considered several publicly available datasets. Here we consider their conclusions in the context of Brawand *et al.* 2011 [@Brawand:2011du], one of the studies they considered. This dataset of gene expression for XX different organs across XX mammal species is the best sampled in their reanalyses. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indiciates no specifificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates his specificity. It is convenient in that it is a single number of defined range for each gene in each species, though of course since the original expression is multidemensional this means much information is discarded. For example, it does not convey which tissue expression is specific to. Homologous genes could each have a tau value of 1, but one gene could have expression that is specific to the kidney and the other to the brain, for example. Their published analyses are based on pairwise comparisons (Figure 1a) betwen tau for each gene family. They found the correlation for orthologs to be significantly greater than the correlation of tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs. From this concluded that their analyses support the ortholog conjecture. They also went one step further - they concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs."

@@

We reanalyzed the their study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [Brawand:2011du]. For each internal node in each gene tree (Figure 1b), we measured the phylogenetically independent contrast (PIC) in the expression across this node [@Felsenstein:1985ua]. This is the difference in values of descendent nodes (*i.e.*, the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches). These contrasts were then annotated by whether they are made across branches that descend from a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of it to look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree is a bit different than the usual application of the method.

@@

To test the ortholog conjecture it is essental to have a null hypothesis that results in a distinct prediction from the ortholog conjecture. A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events (Figure 1b). Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a). Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).

@@

	nodes_contrast %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
        geom_freqpoly( binwidth=.001, position="identity" ) +
        xlab("Phylogenetic independent \ncontrasts (PIC)") +
				xlim(0,0.075) + 
				theme_classic()

@@

The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [Felsenstein:1985ua]. The expected variance is principaly determined by the lengths of the branches leading from the node to these daughters. The same observed difference between daughter nodes will lead to a larger contrast the shorter the branches are to these nodes, as this indicates a greater rate of evolutionary change. The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications can range much wider (Figure XXAscertainment c).

The fact that the expected variance of speciation contrasts is a subset of the range of the expected variance of duplication contrasts could lead to ascertainment biases. We therefore also examine only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range. This reanalysys does not reject the ortholog conjecture either (Figure XXAscertainment d).

@@

In order to better understand why our phylogenetic analysis seem to support a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their specific result. This is imporant since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset. We can replicate their result - we find a higher tau correlation between orthologs (R=`r round( ortholog_r, 3) `) then between paralogs (R=`r round( paralog_r, 3)`).

@@

Paralogs could be more different from each other than ortholgs if they tend to be more distantly related to each other than orthologs are. That would provide more time for differences to accumulate between them, regardless of the history of duplication and speciation.  This is, in fact, what we find in this dataset. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years. This is at least in part because, while the maximum distance between orthogs is two times the depth of the oldest speciation event in the tree, many duplications precede this oldest sampled speciation event. The distance between paralogs resulting from these deeper duplications can be much larger.

@@

The other issue with the pairwise comparisons is that many are summarizing changes along many branches in the phylogeny, not just the branches that follow the divergence event. Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and cnages along all these branches will impact the final pattern. The phylogenetic methods isolate the changes under consideration. There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss. But these affect both methods. The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.

The reason that paralogs pairs have tau values that are more different than orthog pairs is that they tend to be more distantly related, not because there are differences in evolutionary rates of change in tau following duplication and speciation events. This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distintly related the two genes are. This has several implications. One is that it is an example of the limitations of the terms orthology and paralogy [@Dunn:2016er].

@@

While we do demonstrate a common problem with pairwise comparisons in the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that there is not a trait that can be mapped onto the tree, as KMRR's tau can be. A full phylogenetic reanalysis would of course be possible be using upstream analysis products to rederive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.

@@

Some of the most widely used phylogenetic comparative methods [1, 5] are directly applicable to CFG data. There are also unique challenges that will need to be addressed at this new intersection, though. One of the greatest challenges is that most phylogenetic comparative methods have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined (e.g., 2-5). In CFG analyses, there are often far fewer species (e.g., 3-10) because adding taxa is still expensive, but tens of thousands of traits. This creates challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artefacts of project design. We outlined these challenges and potential solutions in the context of gene expression [31].

It is not enough to demonstrate the impact of phylogenetics on CFG analyses, these methods need to be made accessible to the FG community so they can be widely.

Most early phylogenetic comparative methods attempted to remove evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [34]. The application of this newer focus to pCFG provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in FG traits and shifts in rates of evolution of one FG trait following changes in another trait.

@@

Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjectures is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families. This has impotant biological implications. It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as thought. It also has technical implications. Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are. Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs.

The widespread invokation of the ortholog conjecture follows from the expectation that it is strong, widely distributed phenomenon. Our failure to detect a signature consistent with the ortholog conjecture suggests that it may not be, and that it should not be assumed in systems where it has not been explicitly tested.

@@

All files needed to reexecute the anlyses presented in this document are available at https://github.com/caseywdunn/comparative_expression_2017. The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.

@@

The KMRR study [@KryuchkovaMostacci:2016iw] followed excellent practices in reproducibility. They posted all data and code needed to re-execute their analyses at figshare:  https://figshare.com/articles/Tissue-specificity_of_gene_expression_diverges_slowly_between_orthologs_and_rapidly_between_paralogs/3493010/2 . We slightly altered their `Rscript.R` to simplify file paths. This midified script and their data files are available in the gihub respository for this paper, as are the intermediate files that were generated by their analysis script that we used in our own analyses. We also obtained the `Compara.75.protein.nh.emf` gene trees [@Herrero:2016jj] from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and place them in the same directory as this file.

@@

Each of these gene trees includes includes homologous sequences derived from a common ancestral sequence. The tip nodes of a gene tree can include both orthogs and paralogs, and each internal node is an infered speciation or duplication event. Prior to phylogenetic analyses of gene expression, each tree was first pruned to include only tips for which expression data were available. The speciation nodes in the tree were then time calibrated so that branch lengths were consistent across trees. We then calculated the magnitude of the independent contrast for differential expression at each internal node in each of these trees. The independent contrast is the amount of change in expression observed along the branches that arise from the node, scaled by the expected variance [Felsenstein:1985ua]. The expected variance is determined mostly by the length of the branches, but can also be influenced by uncertainty regarding the reconstruction of differential expression at internal nodes. We took the absolute value of these contrasts, since the sign depends only on the arbitrary ordering of the nodes in the trees.

@@

## Ackowledgements

@@

> Supplementary Figure XXAscertainment | Investigation of possible assertainment biases. (a) Magnitude of independent contrasts ploted against node age. Speciation nodes are calibrated to particular times, whereas duplication nodes have a wider range. (b) Density plot of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes. (c) Magnitude of independent contrasts ploted against expected variance, which is largely determined by branch lengths. Contrasts for speciation nodes have a narrower range of expected variance than do contrasts for duplication nodes. (d) Density plot of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes.

>>>>

Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in biology. There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity. Most published comparisons of genome function across species have, however, neglected to use methods that consider the evolutionary relationships among species. This has two potential negative impacts. First, it can lead to the wrong conclusions about the evolution of genome function. Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships. Here we examine two recently published comparative gene expression studies that used multiple pairwise comparison rather than phylogenetic approaches to investigate evolutionary patterns of gene expression. We find problems with the pairwise comparisons used in both studies that undermine their published conclusions. The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpreted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication. Our phylogenetic analysis finds no difference in rates following duplication and speciation. Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs are to each other. The second study claimed to find more evolutionary variance in gene expression at a midpoint of embryonic development than earlier and later when comparing distantly related animals. We instead find that changes along the branch of a single taxon, the ctenophore, were over counted because they were included in each pairwise comparison with this species. These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, including gene expression analyses, and indicates that it will be critical to adopt phylogenetic comparative methods in future work.

@@

The focus of genomics research has quickly shifted from describing genome sequences to functional genomics, the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation. functional genomics, in turn, is now becoming more comparative- there is great interest in understanding how functional genomic variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh]. Comparative Functional Genomics (CFG) analyses are also critical to transferring functional insight across species, and will grow in importance in coming years.

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Chamberlain:2012vx; @Garamszegi:2014ew; @Pagel:1999gc]. These challenges include non-independence of obserbations across species -- more closely related species will tend to be more similar to each other than distantly related species. Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG. Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on multiple pairwise comparisons (Fig. 1A) between contemporary traits that do not account for the fact that evolutionary relationships explain much of the structure of variation across species. This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic approaches can lead to different results in CFG studies. In this manuscript we examine this issue. Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other types of character data [@Chamberlain:2012vx; @Ricklefs:1996wb], this value is apparently not well known to the functional genomics community since the field has until recently largely focused on within-species analyses. Many initial CFG studies have also been focused on only two species at a time, where the differences between phylogenetic and pairwise approaches are not as pronounced. As the number of species considered in each CFG study grows, the shortcomings of pairwise approaches will become far more acute.

Most CFG studies to date compare gene expression across species, often between distinct tissues or developmental time points. Gene expression has been a popular target of CFG studies because, with the advent of RNA-Seq and related transcriptome sequencing methods, it is the most readily collected functional genomics trait across diverse species. Here we reconsider two recent comparative analyses of gene expression data [@KryuchkovaMostacci:2016iw; @Levin:2016bp].

Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw] analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend of have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm]. Using pairwise comparisons (Fig. 1A), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.

The second CFG study we evaluate here is Levin *et al.* [@Levin:2016bp]. This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. They concluded that these animals exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Fig. 2). This contrasts with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw]. Levin *et al.* conclude that this provides biological justification for the concept of phyla and may provide a definition of phyla. We previously described concerns with their interpretations of these results [@Hejnol:2016hw]. Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.

@@

#### Original pairwise analyses

The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is used to relate findings from well-studied genes to related genes that haven't been investigated in detail. It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it can also be considered a hypothesis about the process of evolution. Specifically, it postulates that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

Despite its importance, there have been relatively few tests of the ortholog conjecture. Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz]. Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw]. There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough expression study to date.

The KMRR study considered several publicly available datasets. Here we consider their conclusions in the context of Brawand *et al.* 2011 [@Brawand:2011du], one of the studies they considered. This dataset of gene expression for six organs across 10 species (nine mammals and a bird) is the best sampled in their reanalyses. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue). It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded. For example, it does not convey which tissue expression is specific to. Homologous genes could each have a tau value of 1, but one gene could have expression that is specific to the kidney and the other to the brain, for example.

KMRR's published analyses are based on pairwise comparisons (Figure 1a) between tau within each gene family. They found the correlation of tau for orthologs to be significantly greater than the correlation of tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs. From this they concluded that their analyses support the ortholog conjecture. They also went one step further. They concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".

@@

We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [Brawand:2011du]. For each internal node in each gene tree (Figure 1b), we measured the phylogenetically independent contrast (PIC) in the expression across this node [@Felsenstein:1985ua]. This is the difference in values of descendant nodes (*i.e.*, the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches). These contrasts were then annotated by whether they are made across branches that descend from a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of it to look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree is a bit different than the usual application of the method.

@@

To test the ortholog conjecture it is essential to have a null hypothesis that results in a distinct prediction from the ortholog conjecture. A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events (Figure 1b). Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a). Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).

@@

	nodes_contrast %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
        geom_freqpoly( binwidth=.001, position="identity" ) +
        xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
				ylab( "Density" ) +
				xlim( 0, 0.075 ) + 
				theme_classic()

@@

The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [Felsenstein:1985ua]. The expected variance is principally determined by the lengths of the branches leading from the node to these daughters. The same observed difference between daughter nodes will lead to a larger contrast the shorter the branches are to these nodes, as this indicates a greater rate of evolutionary change. The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications can range much wider (Figure XXAscertainment c).

The fact that the expected variance of speciation contrasts is a subset of the range of the expected variance of duplication contrasts could lead to ascertainment biases. We therefore also examine only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range. This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d).

@@

In order to better understand why our phylogenetic analysis seem to support a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their specific result. This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset. We can replicate their result - we find a higher tau correlation between orthologs (R=`r round( ortholog_r, 3) `) then between paralogs (R=`r round( paralog_r, 3)`).

@@

Paralogs could be more different from each other than orthologs if they tend to be more distantly related to each other than orthologs are. That would provide more time for differences to accumulate between them, regardless of the history of duplication and speciation.  This is, in fact, what we find in this dataset. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years. This is at least in part because, while the maximum distance between orthologs is two times the depth of the oldest speciation event in the tree, many duplications precede this oldest sampled speciation event. The distance between paralogs resulting from these deeper duplications can be much larger.

@@

The other issue with the pairwise comparisons is that many are summarizing changes along many branches in the phylogeny, not just the branches that follow the divergence event. Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pattern. The phylogenetic methods isolate the changes under consideration. There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss. But these affect both methods. The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.

The reason that paralogs pairs have tau values that are more different than ortholog pairs is that they tend to be more distantly related, not because there are differences in evolutionary rates of change in tau following duplication and speciation events. This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are. This has several implications. One is that it is an example of the limitations of the terms orthology and paralogy [@Dunn:2016er]. These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees. Orthology and paralogy are annotations on the tips of the tree that are derived from the infered structure of the tree and annotations of duplication and speciation at internal tree nodes. In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree. Much is lost in the process, though, and for most questions it is much more direct to focus on the structure of the tree and the inferred processes of internal nodes. In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the infered duplication and speciation history on gene phylogenies [@Herrero:2016jj]. These are the same pre-computed trees we use in our phylogenetic re-analyses of their study.

@@

While we do demonstrate a common problem with pairwise comparisons in the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that there is not a trait that can be mapped onto the tree, as KMRR's tau can be. A full phylogenetic reanalysis would of course be possible be using upstream analysis products to re-derive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.

@@

Some of the most widely used phylogenetic comparative methods [1, 5] are directly applicable to CFG data. There are also unique challenges that will need to be addressed at this new intersection, though. One of the greatest challenges is that most phylogenetic comparative methods have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined (e.g., 2-5). In CFG analyses, there are often far fewer species (e.g., 3-10) because adding taxa is still expensive, but tens of thousands of traits. This creates challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design. We outlined these challenges and potential solutions in the context of gene expression [31].

It is not enough to demonstrate the impact of phylogenetic methods on CFG analyses, these methods need to be made accessible to the functional genomics community so they can be widely applied.

Most early phylogenetic comparative methods attempted to remove evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [34]. The application of this newer focus to pCFG provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomics traits and shifts in rates of evolution of one functional genomics trait following changes in another trait.

@@

Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjectures is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families. This has important biological implications. It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as thought. It also has technical implications. Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are. Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs.

The widespread invocation of the ortholog conjecture follows from the expectation that it is strong, widely distributed phenomenon. Our failure to detect a signature consistent with the ortholog conjecture suggests that it may not be, and that it should not be assumed in systems where it has not been explicitly tested.

@@

All files needed to re-execute the analyses presented in this document are available at https://github.com/caseywdunn/comparative_expression_2017. The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.

@@

The KMRR study [@KryuchkovaMostacci:2016iw] followed excellent practices in reproducibility. They posted all data and code needed to re-execute their analyses at figshare:  https://figshare.com/articles/Tissue-specificity_of_gene_expression_diverges_slowly_between_orthologs_and_rapidly_between_paralogs/3493010/2 . We slightly altered their `Rscript.R` to simplify file paths. This modified script and their data files are available in the github repository for this paper, as are the intermediate files that were generated by their analysis script that we used in our own analyses. We also obtained the `Compara.75.protein.nh.emf` gene trees [@Herrero:2016jj] from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and place them in the same directory as this file.

@@

Each of these gene trees includes includes homologous sequences derived from a common ancestral sequence. The tip nodes of a gene tree can include both orthologs and paralogs, and each internal node is an inferred speciation or duplication event. Prior to phylogenetic analyses of gene expression, each tree was first pruned to include only tips for which expression data were available. The speciation nodes in the tree were then time calibrated so that branch lengths were consistent across trees. We then calculated the magnitude of the independent contrast for differential expression at each internal node in each of these trees. The independent contrast is the amount of change in expression observed along the branches that arise from the node, scaled by the expected variance [Felsenstein:1985ua]. The expected variance is determined mostly by the length of the branches, but can also be influenced by uncertainty regarding the reconstruction of differential expression at internal nodes. We took the absolute value of these contrasts, since the sign depends only on the arbitrary ordering of the nodes in the trees.

@@

## Acknowledgments

@@

> Supplementary Figure XXAscertainment | Investigation of possible ascertainment biases. (a) Magnitude of independent contrasts plotted against node age. Speciation nodes are calibrated to particular times, whereas duplication nodes have a wider range. (b) Density plot of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes. (c) Magnitude of independent contrasts plotted against expected variance, which is largely determined by branch lengths. Contrasts for speciation nodes have a narrower range of expected variance than do contrasts for duplication nodes. (d) Density plot of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes.
